Programming Language Survey Paper on Haskell

### 2013-04-22 projects/term_paper/paper.tex
Finished the exception handling section.

>>>>

Haskell, like many other languages, provides ways to utilize error handling code
in the event of abnormal program execution.  However, Haskell is
different from other languages in the sense that the preferred way to handle
these circumstances depends upon the portion of the code which requires action.
To explain this point, functions in Haskell must be classifed as either pure
or impure.  Recall the in previous sections, the Haskell language core was
claimed to be pure functional programming in the sense the functions do
not produce any side-effects beyond their output values.  However, certain
functions in Haskell must break this trend in some respect; such functions include
those responsible for I/O, concurrency, and interfaces to other languages~\cite{haskell-awkward-squad}.
As a side note, this split from the functional paradigm for these types of functions has
been mended by the introduction of monads in Haskell, a topic which unfortunately will not be covered here.

For impure functions, \texttt{exceptions} are the favored way to perform error handling.
The following example provides an error handling function \texttt{handler} associated with the
\texttt{tryThis} function for intervening with I/O file-related issues.
\begin{minted}{haskell}
import System.Environment  
import System.IO  
import System.IO.Error

main = tryThis `catch` handler

tryThis :: IO ()  
tryThis = do (fileName:_) <- getArgs  
    fileContents <- readFile fileName  
    putStrLn $ "File contains " 
        ++ show (length (lines fileContents)) ++ " lines"

handler :: IOError -> IO ()  
handler e 
    -- handle exception
    | isDoesNotExistError e = putStrLn "The file does not exist."  
    -- re-raise exception
    | otherwise = ioError e  
\end{minted}

For pure functions, the Haskell typeclass system was broaded to allow the introduction
of the \texttt{Maybe} algebraic type.  This type extension provides a function a way to
indiciate that it may or may not return a value of the specified type in its declaration
(i.e., it may fail to return any value).  If a function implementing this typeclass
fails this way, the return value will be \texttt{Nothing}.  If it does return a value, say \texttt{x},
that value will be prefaced by the keyword \texttt{Just}.  A pure function illustrating this concept
is given below.
\begin{minted}{haskell}
doubleHead :: [Int] -> Maybe Int
doubleHead [] = Nothing
doubleHead xs = Just ((head xs) * 2)

main = do
    print (doubleHead [2,4,6])  -- output: Just 4
    print (doubleHead [])       -- output: Nothing
\end{minted}
